Transactional Memory Concurrency Verification with Landslide

### 2018-03-15 htm/paper.tex
acks
avl_insert bug!!

\begin{table*}[t]
	\begin{center}
		% TODO: highlight the winning statistic for each test
	\begin{tabular}{cc||r|r|r||r|r|r|r}
			&	&\multicolumn{3}{c||}{\bf Quicksand mode}&\multicolumn{4}{c}{{\bf Maximal state space mode} ({\tt -M})} \\
		\bf buggy test	& \bf params&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}& \ETA{\bf SS size (est.)} \\
		\hline
		\hline
		{\tt htm1}
			& 2,1	&\cpu{  45.78}&\wtm{9.70}&\ints{21}& \cpu{9.47}& \wtm{6.40}& \ints{21	}& \ETA{213} \\
		(assertion)
			& 2,2	&\cpu{  84.14}&\wtm{13.59}&\ints{33}& \cpu{10.39}& \wtm{7.70}& \ints{49	}& \ETA{1536} \\
			& 2,3	&\cpu{ 131.91}&\wtm{20.44}&\ints{73}& \cpu{12.83}& \wtm{9.67}& \ints{113	}& \ETA{10752} \\
			& 2,4	&\cpu{ 255.75}&\wtm{37.56}&\ints{257}& \cpu{18.63}& \wtm{15.86}& \ints{257	}& \ETA{73728} \\
		\cline{2-9}
			& 3,1	&\cpu{ 114.06}&\wtm{17.45}&\ints{15}& \cpu{9.50}& \wtm{6.79}& \ints{21	}& \ETA{13653} \\
			& 3,2	&\cpu{ 109.60}&\wtm{26.16}&\ints{49}& \cpu{10.72}& \wtm{7.97}& \ints{49	}& \ETA{393216} \\
			& 3,3	&\cpu{ 124.80}&\wtm{20.40}&\ints{73}& \cpu{13.84}& \wtm{11.01}& \ints{113	}& \ETA{11010048} \\
			& 3,4	&\cpu{ 227.49}&\wtm{35.15}&\ints{161}& \cpu{31.37}& \wtm{28.53}& \ints{257	}& \ETA{301989888} \\
		\cline{2-9}
			& 4,1	&\cpu{ 53.08}&\wtm{9.79}&\ints{15}& \cpu{9.82}& \wtm{7.00}& \ints{21	}& \ETA{873813} \\
			& 4,2	&\cpu{117.07}&\wtm{19.09}&\ints{33}& \cpu{11.54}& \wtm{8.55}& \ints{49	}& \ETA{100663296} \\
		\hline
		{\tt swapbug}
			& 2,1	&\cpu{ 70.95}&\wtm{13.45}&\ints{16}& \cpu{38.96}& \wtm{13.15}& \ints{109	}& \ETA{194} \\
		(deadlock)
			& 2,2	&\cpu{ 107.28}&\wtm{17.45}&\ints{146}& \cpu{44.73}& \wtm{19.47}& \ints{281	}& \ETA{1620} \\
			& 2,3	&\cpu{ 280.05}&\wtm{38.70}&\ints{352}& \cpu{60.30}& \wtm{35.55}& \ints{718	}& \ETA{12748} \\
			& 2,4	&\cpu{ 617.94}&\wtm{81.50}&\ints{834}& \cpu{108.58}& \wtm{82.60}& \ints{1820	}& \ETA{97823} \\
		\cline{2-9}
			& 3,1	&\cpu{1275.04}&\wtm{163.42}&\ints{771}&\ETA{>30m}&\ETA{>30m}&\ETA{unk.}& \ETA{184984} \\
			& 3,2	&\ETA{>30m}&\ETA{>30m}&\ETA{unk.}&\ETA{>30m}&\ETA{>30m}&\ETA{unk.}& \ETA{3099225} \\
		% whoops, this was with the dumb lock ordering
		%	& 3,1	& 11.19	& 3.46	& 9	& 3.13	& 3.13	& 9	& 432 \\
		%	& 3,2	& 12.41	& 3.76	& 19	& 3.33	& 3.33	& 19	& 5832 \\
		%	& 3,3	& 15.77	& 4.58	& 41	& 3.95	& 3.95	& 41	& 77759 \\
		%	& 3,4	& 24.82	& 6.85	& 89	& 5.20	& 5.20	& 89	& 1026431 \\
		%\cline{2-9}
		%	& 4,1	& 702.33& 98.06	& 361	& 90.94	& 64.27	& 1607	& 138758 \\
		%	& 4,2	&9738.43&1257.20& 3023	& 674.87& 650.58& 16511	& 13755188 \\
	\end{tabular}
	\end{center}
	\caption{Landslide's bug-finding performance on various test configurations.
		Quicksand's workqueue approach optimized for fast bug-finding
		is compared against our maximum-state-space-prioritizing approach for fast verification.
		For each, we list the CPU-time and wall-clock time %(both in seconds)
		elapsed, %until the bug was found,
		plus the number of interleavings of the ultimately buggy state space tested,
		before the bug was found.
		Lastly, Landslide's state space size estimation \cite{estimation},
		though approximate at best,
		confers a sense of the exponential explosion.
	}
	\label{tab:buges}
\end{table*}

@@

\begin{itemize}
	\item {\tt htm1}: The bug from Figure~\ref{fig:htm-example}. %(assertion failure on {\tt count}).
	\item {\tt htm2}: The fixed version as in Figure~\ref{fig:htm-fixed}.
	\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex failure path with a simple {\tt xadd}.
	\item {\tt swap}: Microbenchmark that swaps values in an array.
	\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
	\item {\tt map\_basic}: Concurrent insertion test for the separate-chaining hashmap.
	\item {\tt map\_basicer}: {\tt map\_basic} modified with a larger initial size to skip the resizing step.
		% TODO more
\end{itemize}

@@

Table~\ref{tab:buges}
presents our bug-finding results.
We configured Landslide to run the Quicksand algorithm \cite{quicksand}
% this note obsolete since swapbug was fixed and ends at 3,2
%\footnote{This was necessary in {\tt swapbug}s (4,1) and (4,2) to trigger a switch to the buggy state space.}
shown left,
as well as to prioritize the maximal state space as discussed above,
shown right,
each with a time limit of 30 minutes.
We make two observations/conclusions:
\begin{itemize}
	\item The multiplicative factor in bug-finding speed (2-4x, eyeballing) is generally smaller
		than that of the total number of interleavings (10-100x)
		as the test parameters increase.
		In other words,
		should they exist,
		Landslide find bugs reasonably quickly in these transactional programs
		despite prohibitive exponential explosion in total state space size.
	%\item Despite Quicksand's ability to optimize for finding bugs in overall smaller state spaces,
	%	this does not correlate with better CPU-time or even better wall-clock time.
	%	While possibly a consequence of the modest test size,
	%	this suggests future work should extend the Quicksand algorithm
	%	to prioritize by state space maximality as well as by state space estimate.
	\item Quicksand's ability to find bugs in fewer distinct interleavings %overall smaller state spaces
		does not necessarily correlate with better performance. %CPU- or even wall-clock time.
		%Comparing to the break-even point in Quicksand's evaluation \cite{quicksand},
		%most of these test configurations fall on the side favouring the single-state-space approach,
		Most of our tests are too small for its approach to pay off,
		with {\tt swapbug}(3,1) as its lone win.
		While the prior work showed plenty more wins \cite{quicksand},
		%an upper bound also exists:
		%in {\tt swapbug}(3,2),
		%not even the {\em minimal} state space was completed in time.
		%this suggests
		future MCs could prioritize state spaces using not just size estimation
		but state space maximality as well
		%but a hybrid approach also conisdering state space maximality and preemption bounds \cite{chess-icb},
		to soften the trade-off
		both for smaller tests and for verification.
\end{itemize}
Speaking of which...

@@

Landslide proved the following tests correct using {\tt -M} mode.
With no bugs to find, comparing the different testing modes against each other is meaningless;
we simply present their state space sizes and runtime in Table~\ref{tab:verifs}.

\newcommand\ETAdag[1]{\ETA{\ensuremath{\dagger}#1}}
\begin{table}[h]
	\begin{center}
		\begin{tabular}{cc||r|r}
			\bf test & \bf params & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\hline
			\hline
			{\tt counter}
			& (2,1) & \cpu{5.57}	& \ints{30}	\\
			& (2,2) & \cpu{15.53}	& \ints{384}	\\
			& (2,3) &\cpu{20.00}	& \ints{5280}	\\
			& (2,4) &\cpu{2211.10}	& \ints{75264}	\\
			& (3,1) & \cpu{57.90}	& \ints{1960}	\\
			& (3,2) &		& 	\\
			\hline
			{\tt htm2}
			& (2,1) & \cpu{18.57}	& \ints{294}	\\
			& (2,2) & \cpu{133.78}	& \ints{4902}	\\
			& (2,3) &\cpu{1986.98}	& \ints{79017}	\\
			& (3,1) &\cpu{11672.15}	& \ints{467730}	\\
			& (3,2) &	&	\\
			\hline
			{\tt swap}
			& (2,1) & \cpu{38.77}	& \ints{228}	\\
			& (2,2) &	&	\\
			& (3,1) &	&	\\ % running atm
			\hline
			{\tt map\_basic}
			& (2,1) & \ETAdag{10d 16h} & \ETAdag{16388977} \\
			%[JOB 15] progress: 70330/16388977 brs (0.429130%), ETA 10d 16h 34m 15s (elapsed 43m 21s)
			%[JOB 15] TIMED OUT (0.429130%; ETA 10d 16h 34m 15s)
			\hline
			{\tt map\_basicer}
			& (2,1) & \cpu{877.44}& \ints{28635}	\\
			& (2,2) & \ETAdag{2d 6h} & \ETAdag{5925634} \\
			%[JOB 6] progress: 35042/5925634 brs (0.591363%), ETA 2d 6h 40m 30s (elapsed 21m 12s)
			%[JOB 6] TIMED OUT (0.591363%; ETA 2d 6h 40m 30s)
		\end{tabular}
	\end{center}
	\caption{Transactional tests verified (or not) by Landslide.}
	\label{tab:verifs}
\end{table}

Several larger tests we were unable to complete before the submission deadline are also shown.
%each timing out after 1 hour. % TODO: ??
%Their completion times are listed as {\em unk.}
These are indicated with \ETAdag{}
and their listed ETAs and state space sizes represent Landslide's estimate %at the moment of timeout.
after a timeout of 1 hour
(hopefully enough for data-race PPs to saturate and estimates to stablize,
although note the two estimate types use different algorithms so may disagree significantly).
%For these, Table~\ref{tab:timeouts} presents their size estimates at the moment of timeout.
Future work may expand our coverage upon this frontier \cite{landslide-phdthesis}.
%\begin{table}[h]
%	\begin{center}
%		\begin{tabular}{cc|r}
%			test & params & SS size (est.) \\
%			\hline
%		\end{tabular}
%	\end{center}
%	\caption{Transactional test verification timeouts.}
%	\label{tab:timeouts}
%\end{table}

@@

\begin{table}[h]
	\begin{center}
	\begin{tabular}{l|c|c|c}
		\bf load & \cpu{\bf cpu (s)} & \bf vs self avg & \bf vs baseline \\
		\hline
		{\sf none	} & \cpu{14.95 $\pm$ 0.17} & 0.99-1.02x	& (baseline)	\\
		{\sf vid-L	} & \cpu{15.13 $\pm$ 0.10} & 0.99-1.01x	& 1.01x	\\
		{\sf ff/c	} & \cpu{15.55 $\pm$ 0.16} & 0.99-1.02x	& 1.04x	\\
		{\sf sm5	} & \cpu{16.63 $\pm$ 0.07} & 0.99-1.01x	& 1.11x \\
		{\sf ff/c+vid-S	} & \cpu{17.09 $\pm$ 0.34} & 0.98-1.05x	& 1.14x	\\
		{\sf ls		} & \cpu{19.11 $\pm$ 0.32} & 0.97-1.03x	& 1.28x	\\
		% simply love title screen
		%{\sf sm5+sl}	& 19.18 $\pm$ 0.32	& 0.97-1.02x	& 1.28x \\
		{\sf ff/c+ls	} & \cpu{19.66 $\pm$ 0.50} & 0.96-1.02x	& 1.31x	\\
	\end{tabular}
	\end{center}
	\caption{Performance variance on {\tt counter}(2,2) with other programs running on the test machine.
		{\sf vid-L} is full-screen video (played locally with {\tt mplayer}),
		{\sf ff/c} is Firefox and Chrome (idle, $\le$20 tabs),
		{\sf sm5} is StepMania 5.1 (during gameplay \cite{itg2}),
		{\sf vid-S} is full-screen video (streamed via Crunchyroll),
		{\sf ls} is a 2nd instance of Landslide.
		Average of 10 samples, $\pm N$ is 1 stddev.
	}
	\label{tab:variants}
\end{table}

Table~\ref{tab:variants2} shows
the impact of Landslide needing to automatically annotate a test being run for the first time,
rather than reusing existing instrumentation,
as well as the variance of Quicksand mode (i.e., without {\tt -M}).
%(2128 interleavings total).
The {\sf quicksand} variance was surprisingly bimodal,
with 6 samples in a distribution of 106.14$\pm$1.22 and 4 in 142.11$\pm$1.67,
suggesting two distinct scheduling patterns for its workqueue threads.
Future work should figure out why.

\begin{table}[h]
	\begin{center}
	\begin{tabular}{l|r|c}
		\bf Landslide mode & \bf total int's & \cpu{\bf cpu (s)} \\
		\hline
		{\sf verif} ({\tt -M})		& 403  & \cpu{14.95 $\pm$ 00.17} \\
		{\sf reinstrument}		& 403  & \cpu{24.72 $\pm$ 00.10}	\\
		{\sf quicksand} (no {\tt -M})	& 2128 & \cpu{120.53 $\pm$ 18.62}	\\
	\end{tabular}
	\end{center}
	\caption{Performance variance on {\tt counter}(2,2) in various modes.
		Average of 10 samples, $\pm N$ is 1 stddev.
	}
	\label{tab:variants2}
\end{table}

@@

Outside of Table~\ref{tab:variants}, we fixed our environment by measuring all performance numbers
with Firefox and Chrome as the only other significant machine load.
%(we gotta stay amused somehow).
We believe the exponential differences among completion times justifies the absence of error bars,
which would be expected to show 2\% variance if extrapolating from this experiment.
The numbers of interleavings in each state space are, of course, deterministic and do not vary across runs.

@@

Many such branches may be equivalent;
for example,
explicit and conflict aborts need not be tested separately in transactions
whose failure paths do not distinguish the cause,
and the perennial {\tt \_XABORT\_RETRY} abort can be skipped entirely
if the client abstracts it away behind a retry loop.
In fact, applying both reductions simultaneously
produces a state space smaller than the original;
testing them by hand (trusting their soundness by visual inspection)
on {\tt map\_basicer}(2,1) reduced its 28635 interleavings to 11577
(although {\tt map\_basic}(2,1) lies still beyond reach).
%We tested both reductions by hand, trusting their soundness by visual inspection,
%on {\tt map\_basic}(2,1),
%and preview the resulting state space sizes in Table~\ref{tab:abort-codes}.
%
%\begin{table}[h]
%	\caption{...}
%	\label{tab:abort-codes}
%\end{table}
%
Future work could use static or dynamic flow analysis to
identify such reduction opportunities automatically;
for now, this feature is disabled by default but accessible via the {\tt -A} command-line option
to Landslide (in addition to {\tt -X}).

@@

>>>>

\begin{table*}[t]
	\begin{center}
	\begin{tabular}{cc||r|r|r||r|r|r|r}
			&	&\multicolumn{3}{c||}{\bf Quicksand mode}&\multicolumn{4}{c}{{\bf Maximal state space mode} ({\tt -M})} \\
		\bf buggy test	& \bf params&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}& \ETA{\bf SS size (est.)} \\
		\hline
		\hline
		{\tt htm1}
			& 2,1	&\cpu{  45.78}&\wtm{9.70}&\ints{21}& \cpu{9.47}& \wtm{6.40}& \ints{21	}& \ETA{213} \\
		(assertion)
			& 2,2	&\cpu{  84.14}&\wtm{13.59}&\ints{33}& \cpu{10.39}& \wtm{7.70}& \ints{49	}& \ETA{1536} \\
			& 2,3	&\cpu{ 131.91}&\wtm{20.44}&\ints{73}& \cpu{12.83}& \wtm{9.67}& \ints{113	}& \ETA{10752} \\
			& 2,4	&\cpu{ 255.75}&\wtm{37.56}&\ints{257}& \cpu{18.63}& \wtm{15.86}& \ints{257	}& \ETA{73728} \\
		\cline{2-9}
			& 3,1	&\cpu{ 114.06}&\wtm{17.45}&\ints{15}& \cpu{9.50}& \wtm{6.79}& \ints{21	}& \ETA{13653} \\
			& 3,2	&\cpu{ 109.60}&\wtm{26.16}&\ints{49}& \cpu{10.72}& \wtm{7.97}& \ints{49	}& \ETA{393216} \\
			& 3,3	&\cpu{ 124.80}&\wtm{20.40}&\ints{73}& \cpu{13.84}& \wtm{11.01}& \ints{113	}& \ETA{11010048} \\
			& 3,4	&\cpu{ 227.49}&\wtm{35.15}&\ints{161}& \cpu{31.37}& \wtm{28.53}& \ints{257	}& \ETA{301989888} \\
		\cline{2-9}
			& 4,1	&\cpu{ 53.08}&\wtm{9.79}&\ints{15}& \cpu{9.82}& \wtm{7.00}& \ints{21	}& \ETA{873813} \\
			& 4,2	&\cpu{117.07}&\wtm{19.09}&\ints{33}& \cpu{11.54}& \wtm{8.55}& \ints{49	}& \ETA{100663296} \\
		\hline
		{\tt swapbug}
			& 2,1	&\cpu{ 70.95}&\wtm{13.45}&\ints{16}& \cpu{38.96}& \wtm{13.15}& \ints{109	}& \ETA{194} \\
		(deadlock)
			& 2,2	&\cpu{ 107.28}&\wtm{17.45}&\ints{146}& \cpu{44.73}& \wtm{19.47}& \ints{281	}& \ETA{1620} \\
			& 2,3	&\cpu{ 280.05}&\wtm{38.70}&\ints{352}& \cpu{60.30}& \wtm{35.55}& \ints{718	}& \ETA{12748} \\
			& 2,4	&\cpu{ 617.94}&\wtm{81.50}&\ints{834}& \cpu{108.58}& \wtm{82.60}& \ints{1820	}& \ETA{97823} \\
		\cline{2-9}
			& 3,1	&\cpu{1275.04}&\wtm{163.42}&\ints{771}&\ETA{>30m}&\ETA{>30m}&\ETA{unk.}& \ETA{184984} \\
			& 3,2	&\ETA{>30m}&\ETA{>30m}&\ETA{unk.}&\ETA{>30m}&\ETA{>30m}&\ETA{unk.}& \ETA{3099225} \\
		\hline
		{\tt avl\_insert}
			& 2,2	&\cpu{488.07}&\wtm{64.77}&\ints{83}	&\cpu{81.00}&\wtm{40.30}&\ints{336}&\ETA{379982} \\
		(see text)
			% XXX: this 2,3 -M-mode bug was found *mostly* running in isolation,
			% but for the last few minutes i had another landslide running in parallel bc i thought itd time out.
			% according to variance expt, it should be rerun for a *minor* speedup, but i can't be bothered
			& 2,3	&\cpu{2670.87}&\wtm{330.45}&\ints{3066}	&\cpu{1331.79}&\wtm{1274.36}&\ints{13926}&\ETA{96248131} \\
		\cline{2-9}
			& 3,1	&\cpu{222.02}&\wtm{40.04}&\ints{28}	&\cpu{69.99}&\wtm{24.25}&\ints{78}&\ETA{1572107} \\
		% whoops, this was with the dumb lock ordering
		%	& 3,1	& 11.19	& 3.46	& 9	& 3.13	& 3.13	& 9	& 432 \\
		%	& 3,2	& 12.41	& 3.76	& 19	& 3.33	& 3.33	& 19	& 5832 \\
		%	& 3,3	& 15.77	& 4.58	& 41	& 3.95	& 3.95	& 41	& 77759 \\
		%	& 3,4	& 24.82	& 6.85	& 89	& 5.20	& 5.20	& 89	& 1026431 \\
		%\cline{2-9}
		%	& 4,1	& 702.33& 98.06	& 361	& 90.94	& 64.27	& 1607	& 138758 \\
		%	& 4,2	&9738.43&1257.20& 3023	& 674.87& 650.58& 16511	& 13755188 \\
	\end{tabular}
	\end{center}
	\caption{Landslide's bug-finding performance on various test configurations.
		Quicksand's workqueue approach optimized for fast bug-finding
		is compared against our maximum-state-space-prioritizing approach for fast verification.
		For each, we list the CPU-time and wall-clock time %(both in seconds)
		elapsed, %until the bug was found,
		plus the number of interleavings of the ultimately buggy state space tested,
		before the bug was found.
		Lastly, Landslide's state space size estimation \cite{estimation},
		though approximate at best,
		confers a sense of the exponential explosion.
	}
	\label{tab:buges}
\end{table*}

@@

\begin{itemize}
	\item {\tt htm1}: The bug from Figure~\ref{fig:htm-example}. %(assertion failure on {\tt count}).
	\item {\tt htm2}: The fixed version as in Figure~\ref{fig:htm-fixed}.
	\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex failure path with a simple {\tt xadd}.
	\item {\tt swap}: Microbenchmark that swaps values in an array.
	\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
	\item {\tt avl\_insert}: AVL tree concurrent insertion test.
	\item {\tt avl\_fixed}: {\tt avl\_insert} with the AVL bug fixed. %(spoilers!!)
	\item {\tt map\_basic}: Separate-chaining hashmap concurrent insertion test.
	\item {\tt map\_basicer}: {\tt map\_basic} modified with a larger initial size to skip the resizing step.
\end{itemize}

@@

Table~\ref{tab:buges}
presents our bug-finding results.
We configured Landslide to run the Quicksand algorithm \cite{quicksand}
% this note obsolete since swapbug was fixed and ends at 3,2
%\footnote{This was necessary in {\tt swapbug}s (4,1) and (4,2) to trigger a switch to the buggy state space.}
shown left,
as well as to prioritize the maximal state space as discussed above,
shown right,
each with a time limit of 30 minutes.
We draw three main conclusions from this data.
%\begin{itemize}
	%\item

		{\bf Finding bugs quickly.}
		The multiplicative factor in bug-finding speed (2-4x, eyeballing) is generally smaller
		than that of the total number of interleavings (10-100x)
		as the test parameters increase.
		In other words,
		should they exist,
		Landslide find bugs reasonably quickly in these transactional programs
		despite prohibitive exponential explosion in total state space size.
		This corroborates the prior work \cite{quicksand},
		extending its good news to the world of HTM.
	%\item

		{\bf New bugs.}
		In addition to the bugs we intentionally wrote in {\tt htm1} and {\tt swapbug},
		to our pleasant surprise
		Landslide also found a previously-unknown bug in the transactional AVL tree,
		exposed by {\tt avl\_insert} with any parameters higher than (2,1).
		Figure~\ref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
		% landslide-trace-1521093835.18927.html
		This manifested
		alternately
		%in (2,2) and (3,1)
		as a segfault
		% landslide-trace-1521092404.23772.html
		and
		%in (2,3)
		as a consistency-check assertion failure.
		The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
		in practice (between it and {\tt \_xbegin()}),
		whereas MC is blind to such matters of chance.%
		\footnote{Considering the loop does not affect the test's possible behaviours,
		only its {\em likely} ones,
		we removed it in Landslide's version of the test to keep the state space size manageable.}
		%We removed the loop in Landslide's version of the test
		%%To make this test Landslide-friendly, we removed that loop,
		%which reduces the state space size, but maintains the program's semantics.}
		Moreover, we conclude that even Figure~\ref{fig:htm-fixed}'s protocol's very proposer % こそさえ
		getting it wrong motivates the need for MC on such programs,
		and perhaps also means TSX primitives should be used only behind higher-level abstractions
		such as lock elision \cite{lock-elision}.

% TODO syntax hilighting
\begin{figure}[t]
	\begin{center}
		\begin{tabular}{l}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		\texttt{while (\_retry);} \\
		\texttt{if (\_xbegin() == SUCCESS) \{} \\
		\texttt{~~~~tie(\_root,inserted) = \_insert(\_root,n); } \\
		\texttt{~~~~\_xend();} \\
		\texttt{\} else \{} \\
		\texttt{~~~~pthread\_mutex\_lock(\&\_tree\_lock);} \\
		\texttt{~~~~\_retry = true;} \\
		\texttt{~~~~tie(\_root,inserted) = \_insert(\_root,n); } \\
		\texttt{~~~~\_retry = false;} \\
		\texttt{~~~~pthread\_mutex\_unlock(\&\_tree\_lock);} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	%\caption{mario man had a very large house. or a castle.}
	\caption{Unmodified code from {\tt htmavl.cpp} with the previously-unknown bug found by Landslide.
		The transaction path fails to check {\tt \_retry},
		leading to data races and corruption just as in {\tt htm1}.
		}
	\label{fig:avlbug}
\end{figure}
	%\item Despite Quicksand's ability to optimize for finding bugs in overall smaller state spaces,
	%	this does not correlate with better CPU-time or even better wall-clock time.
	%	While possibly a consequence of the modest test size,
	%	this suggests future work should extend the Quicksand algorithm
	%	to prioritize by state space maximality as well as by state space estimate.
	%\item

		{\bf Performance comparison.}
		Quicksand's ability to find bugs in fewer distinct interleavings %overall smaller state spaces
		does not necessarily correlate with better performance. %CPU- or even wall-clock time.
		%Comparing to the break-even point in Quicksand's evaluation \cite{quicksand},
		%most of these test configurations fall on the side favouring the single-state-space approach,
		Most of our tests are too small for its approach to pay off,
		with {\tt swapbug}(3,1) and {\tt avl\_insert}(2,3) as its notable wins.
		While the prior work showed plenty more wins \cite{quicksand},
		%an upper bound also exists:
		%in {\tt swapbug}(3,2),
		%not even the {\em minimal} state space was completed in time.
		%this suggests
		future MCs could prioritize state spaces using not just size estimation
		but state space maximality as well
		%but a hybrid approach also conisdering state space maximality and preemption bounds \cite{chess-icb},
		to soften the trade-off
		both for smaller tests and for verification.
%\end{itemize}
Speaking of which...

@@

Landslide proved the following tests correct.
With no bugs to find, comparing the different testing modes against each other is meaningless;
we simply present their state space sizes and runtime (using {\tt -M}) in Table~\ref{tab:verifs}.

\newcommand\ETAdag[1]{\ETA{\ensuremath{\dagger}#1}}
\begin{table}[h]
	\begin{center}
		\begin{tabular}{cc||r|r}
			\bf test & \bf params & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\hline
			\hline
			{\tt counter}
			& 2,1 & \cpu{5.57}	& \ints{30}	\\
			& 2,2 & \cpu{15.53}	& \ints{384}	\\
			& 2,3 &\cpu{20.00}	& \ints{5280}	\\
			& 2,4 &\cpu{2211.10}	& \ints{75264}	\\
			& 3,1 & \cpu{57.90}	& \ints{1960}	\\
			%& 3,2 &		& 	\\
			\hline
			{\tt htm2}
			& 2,1 & \cpu{18.57}	& \ints{294}	\\
			& 2,2 & \cpu{133.78}	& \ints{4902}	\\
			& 2,3 &\cpu{1986.98}	& \ints{79017}	\\
			& 3,1 &\cpu{11672.15}	& \ints{467730}	\\
			%& 3,2 &	&	\\
			\hline
			{\tt swap}
			& 2,1 & \cpu{38.77}	& \ints{228}	\\
			& 2,2 &	&	\\ % TODO: fix, or use timeout if you HAVE to
			%[JOB 8] progress: 97913/161897 brs (60.478362%), ETA 1h 29m 30s (elapsed 59m 52s)
			% [JOB 8] TIMED OUT (60.478362%; ETA 1h 29m 30s)

			& 3,1 & \ETAdag{8d 11h} & \ETAdag{411312}	\\
			%[JOB 8] progress: 103873/411312 brs (25.254021%), ETA 8d 11h 36m 0s (elapsed 59m 42s)
			%[JOB 8] TIMED OUT (25.254021%; ETA 8d 11h 36m 0s)
			\hline
			{\tt avl\_insert}
			& 2,1 & \cpu{1083.35}	& \ints{40062}	\\
			%& 2,2 & \multicolumn{2}{c}{\em (bug!)} \\
			\hline
			{\tt map\_basic}
			& 2,1 & \ETAdag{10d 16h} & \ETAdag{16388977} \\
			%[JOB 15] progress: 70330/16388977 brs (0.429130%), ETA 10d 16h 34m 15s (elapsed 43m 21s)
			%[JOB 15] TIMED OUT (0.429130%; ETA 10d 16h 34m 15s)
			\hline
			{\tt map\_basicer}
			& 2,1 & \cpu{877.44}& \ints{28635}	\\
			& 2,2 & \ETAdag{2d 6h} & \ETAdag{5925634} \\
			%[JOB 6] progress: 35042/5925634 brs (0.591363%), ETA 2d 6h 40m 30s (elapsed 21m 12s)
			%[JOB 6] TIMED OUT (0.591363%; ETA 2d 6h 40m 30s)
		\end{tabular}
	\end{center}
	\caption{Transactional tests verified (or not) by Landslide.}
	\label{tab:verifs}
\end{table}

Several larger tests we were unable to complete before the submission deadline are also shown.
%Their completion times are listed as {\em unk.}
These are indicated with $\dagger$ %\ETAdag{}
and their listed ETAs and state space sizes represent Landslide's estimate %at the moment of timeout.
after a timeout of 1 hour
(hopefully enough for data-race PPs to saturate and estimates to stablize,
although note the two estimate types use different algorithms so may disagree significantly).
%For these, Table~\ref{tab:timeouts} presents their size estimates at the moment of timeout.
Future work may expand our coverage upon this frontier \cite{landslide-phdthesis}.
%\begin{table}[h]
%	\begin{center}
%		\begin{tabular}{cc|r}
%			test & params & SS size (est.) \\
%			\hline
%		\end{tabular}
%	\end{center}
%	\caption{Transactional test verification timeouts.}
%	\label{tab:timeouts}
%\end{table}

@@

\begin{table}[t]
	\begin{center}
	\begin{tabular}{l|c|c|c}
		\bf load & \cpu{\bf cpu (s)} & \bf vs self avg & \bf vs baseline \\
		\hline
		{\sf none	} & \cpu{14.95 $\pm$ 0.17} & 0.99-1.02x	& (baseline)	\\
		{\sf vid-L	} & \cpu{15.13 $\pm$ 0.10} & 0.99-1.01x	& 1.01x	\\
		{\sf ff/c	} & \cpu{15.55 $\pm$ 0.16} & 0.99-1.02x	& 1.04x	\\
		{\sf sm5	} & \cpu{16.63 $\pm$ 0.07} & 0.99-1.01x	& 1.11x \\
		{\sf ff/c+vid-S	} & \cpu{17.09 $\pm$ 0.34} & 0.98-1.05x	& 1.14x	\\
		{\sf ls		} & \cpu{19.11 $\pm$ 0.32} & 0.97-1.03x	& 1.28x	\\
		% simply love title screen
		%{\sf sm5+sl}	& 19.18 $\pm$ 0.32	& 0.97-1.02x	& 1.28x \\
		{\sf ff/c+ls	} & \cpu{19.66 $\pm$ 0.50} & 0.96-1.02x	& 1.31x	\\
	\end{tabular}
	\end{center}
	\caption{Performance variance on {\tt counter}(2,2) with other programs running on the test machine.
		{\sf vid-L} is full-screen video (played locally with {\tt mplayer}),
		{\sf ff/c} is Firefox and Chrome (idle, $\le$20 tabs),
		{\sf sm5} is StepMania 5.1 (during gameplay \cite{itg2}),
		{\sf vid-S} is full-screen video (streamed via Crunchyroll),
		{\sf ls} is a 2nd instance of Landslide.
		Average of 10 samples, $\pm N$ is 1 stddev.
	}
	\label{tab:variants}
\end{table}

Table~\ref{tab:variants2} shows
the impact of Landslide needing to automatically annotate {\tt counter}(2,2) being run for the first time,
rather than reusing existing instrumentation,
as well as the variance of Quicksand mode. %(i.e., without {\tt -M}).
%(2128 interleavings total).
The {\sf quicksand} variance was surprisingly bimodal,
with 6 samples in a distribution of 106.14$\pm$1.22 and 4 in 142.11$\pm$1.67,
suggesting two distinct scheduling patterns for its workqueue threads.
Future work should figure out why.

\begin{table}[t]
	\begin{center}
	\begin{tabular}{l|r|c}
		\bf Landslide mode & \bf total int's & \cpu{\bf cpu (s)} \\
		\hline
		{\sf verif} ({\tt -M})		& 403  & \cpu{~~14.95 $\pm$ 00.17} \\
		{\sf reinstrument}		& 403  & \cpu{~~24.72 $\pm$ 00.10}	\\
		{\sf quicksand} (no {\tt -M})	& 2128 & \cpu{120.53 $\pm$ 18.62}	\\
	\end{tabular}
	\end{center}
	\caption{Performance variance on {\tt counter}(2,2) in various modes.
		Average of 10 samples, $\pm N$ is 1 stddev.
	}
	\label{tab:variants2}
\end{table}

@@

Outside of this experiment, we fixed our environment by measuring all performance numbers
with Firefox and Chrome as the only other significant machine load.
%(we gotta stay amused somehow).
We believe the exponential differences among completion times justifies the absence of error bars,
which would be expected to show 2\% variance if extrapolating from this experiment.
The numbers of interleavings in each state space are, of course, deterministic and do not vary across runs.

@@

Many such branches may be equivalent;
for example,
explicit and conflict aborts need not be tested separately in transactions
whose failure paths do not distinguish the cause,
and the perennial {\tt \_XABORT\_RETRY} abort can be skipped entirely
if the client abstracts it away behind a retry loop.%
\footnote{Proof left to future work. We've proved enough here already.}
In fact, applying both reductions simultaneously
produces a state space smaller than the original;
testing them by hand
on {\tt map\_basicer}(2,1)%
\footnote{Trusting the reductions' soundness from visual inspection of the test.}
reduced its 28635 interleavings to 11577
(although {\tt map\_basic}(2,1) lies still beyond reach).
%We tested both reductions by hand, trusting their soundness by visual inspection,
%on {\tt map\_basic}(2,1),
%and preview the resulting state space sizes in Table~\ref{tab:abort-codes}.
%
%\begin{table}[h]
%	\caption{...}
%	\label{tab:abort-codes}
%\end{table}
%
Future work could use static or dynamic flow analysis to
identify such reduction opportunities automatically;
for now, this feature is disabled by default but accessible via the {\tt -A} command-line option
to Landslide (in addition to {\tt -X}).

@@

\acks
